### 2015-04-07 paper/chemokine_main_revised.tex
RSI revisions

Identifying the critical components of this complex process is key to understanding viral pathogenicity and designing therapeutic interventions.  However, determining the relative contribution of these myriad experimental components is challenging. We address this challenge with computational modeling, which gives us the advantage of studying relative contributions of different aspects of the host immune response and accounts for strain-specific viral dynamics in the model.

Rather than developing a comprehensive immune system model we focus on one aspect of the adaptive response specifically, studying how activated antigen-specific CD8 T cells migrate from the lymph node to the focus of infection (FOI) in the human lung.  We assume that chemotactic gradients play an important role in guiding T cells to the FOI and that T cells resist extravasation and recirculate in the absence of inflammatory signals.  Chemotactic proteins are known to enhance T cell recruitment in both acute infections and chronic inflammatory diseases \cite{Bromley2008, Medoff2005, Castellino2006 ,Gunn1998, Okada2005}, where recruited T cells mediate either protection or immunopathology.  Infected epithelial cells secrete chemokines \cite{Chan2005}, and contact with CD8 T cells also triggers secretion \cite{Zhao2000}. Activated CD8 T effector cells upregulate chemotactic protein receptors, particularly CXCR3 and CCR5 \cite{Hoji2005, Groom2011a}, responding to their primary ligands CXCL10 (IP-10) and CCL5 (RANTES) respectively.  Previous work showed that efficient recruitment of T cells to the FOI is crucial for the eventual clearance of the virus \cite{Cerwenka1999, Kim2011}.  We extend this work to examine how the unique properties of the different strains of influenza may affect this process.

Here, we present an agent-based model (ABM) of the secondary immune response to an influenza infection in the lung.  Previous models have shed light on the details of the primary and secondary immune response, some by applying ordinary differential equation (ODE) models to empirical data \cite{Miao2010, Saenz2010, Handel2008, Lee2009, Murillo2013}, and some by using spatial models to examine the interactions between dendritic cells and T cells in the lymph node \cite{Beltman2007, Zheng2008, Vroomans2012, Textor2014}.  We parameterize our model with \textit{in vitro} generated measurements of viral expansion dynamics as well as published data of host immune responses to flu infection.  The model was used to determine the pathogenic potential of the three individual flu strains: seasonal H1N1, avian H5N1, and pandemic H1N1.  These three strains were selected for their differences in \textit{in vitro} replication rates \cite{Mitchell2011} as well as differences in \textit{in vivo} severity of human and animal pneumonia and mortality.  The model describes the dynamic, spatially explicit evolution of viral infections in the context of the developing CD8 T cell response.  Sensitivity analysis identifies parameters that are candidates for explaining the dramatic observed differences in pathogenicity among these different flu strains.

@@

The spatial nature of the model identifies physical constraints on T cells' ability to clear the rapidly replicating pH1N1 infection.  Because T cells in the model use the chemokine gradient to locate the FOI, they tend to cluster in areas of high chemokine concentration.  The model suggests that the chemokine response of the host cells may actually hinder T cells' ability to find and clear infected cells in the presence of the rapidly replicating pandemic strain.  Finally, the sensitivity analysis shows the isolated effects of each parameter in the model.  Based on this analysis, each parameter was classified as sensitive or not.  We find that most parameters can be individually varied across a broad range of values without strong effect on predicted outcomes.  Further, the parameter classification is consistent across each broad parameter type. Specifically, chemokine-related parameters are stable within biologically plausible ranges.  In contrast, parameters related to viral kinetics have a strong effect on model results.  We conclude that influenza infection dynamics are dominated by the replication properties of the virus and spatial properties of the infection, confirming that the higher replication rate of 2009 pH1N1 influenza obscured contributions of the chemotactic signals, leading to uncontrolled infection \textit{in silico}.

@@

The model focuses on activated CD8 T cells migrating through the vascular and lymph networks, including their movement over tissue after extravasation (Fig.~\ref{fig:systemchart}).  The lung is represented as a two-dimensional sheet of healthy epithelial cells (S1.3).   Because we assume activated T cells descend at random into branches of the vascular network, T cells in the model are introduced uniformly at random across the modeled lung surface.  If cytokine signal is detected on the local endothelium the T cell remains in tissue and follows the chemotactic gradient to the FOI.  T cells that do not encounter cytokine recirculate to the lymph node where they reenter the bronchial vascular network.  Once inside the endothelium, T cells follow the chemokine gradient to localized areas of maximum chemokine concentration.  When a T cell contacts an infected epithelial cell it induces apoptosis (S1.4).  Otherwise, infected cells continue to secrete virus for a fixed time and then die.

@@

Our model captures important aspects of T cell response to influenza infection which have not been addressed by earlier T cell models.  The results of our sensitivity analysis are important for future biological modeling and experimental design.  Empirical verification of the model's sensitive parameters (T cell production rate, viral response to IgM, infectivity, viral decay rate, viral diffusion rate, and viral production rate) will be valuable to future studies.  Finally, both qualitative and quantitative analysis of our model suggests therapeutic interventions of influenza infection in the lung will likely be most effective by focusing on viral control rather than enhancing T cell responses.

@@

\begin{table}[!ht]
\begin{center}
\begin{tabular}{| l l l |}
  \hline                        
  Referenced Parameters & Value & Source \\
  \hline
  Viral Diffusion in Airway & .0318 $\mu$m$^2$/s & \cite{Beauchemin2006} \\
  Viral Decay in Airway &  1/day & \cite{Lee2009} \\
  Chemokine Diffusion Rate & .318 $\mu$m$^2$/s & \cite{Beauchemin2006} \\
  Incubation Time &  10 hours & \cite{Mitchell2011} \\
  Epithelial Cell Radius & 5 $\mu$m & \cite{Elbert1999} \\
  T Cell Radius & 5 $\mu$m & \cite{abbas2011cellular} \\
  T Cell Production Rate & 1257 cells/hour & \cite{Miao2010} \\ 
  T Cell Speed & 6 $\mu$m/min & \cite{Egen2011} \\
  Blood Circulation Time & 6 seconds & \cite{Banerjee2010b} \\
  T Cell Sensitivity to Chemokine & 100 ng/mL & \cite{Nandagopal2011} \\
  Onset of ATC$^\ddagger$ Lymph Node Exit & Day 5 & \cite{Banerjee2011} \\
  Viral Response to IgM & Viral decay of 10/day & \cite{Diamond2003} \\
  IgM Onset & Day 4 & \cite{Diamond2003} \\
  \hline
  \hline                        
  Estimated Parameters & Value & Footnote \\
  \hline
  Chemokine Decay Rate & $3.8508\cdot10^{-4}$/s & 1\\
  Infectivity & 0.5 cell/hour$\cdot$virion &  2 \\
  Expression Time & 16.7 hours & \cite{Mitchell2011}$^3$ \\
  T Cell Expected Kill Time & 10 min & 4 \\
  Apoptosis Time & 60 minutes & \cite{Ganusov2008}$^5$ \\
  T Cell Age (at FOI$^\ast$) & 2 hours & 6 \\
  T Cell Age (in Blood) & 3 days & 6 \\
  \hline  
\end{tabular}
\caption{Values are taken from experimental literature if possible and from earlier modeling papers if not.  If we could not find a value in any literature we estimated parameters as follows.  1) Corresponds to a 30 minute half-life.  2) Epithelial cells are infected at a probabilistic rate such that the expected time for infection in the presence of a single virion is 2 hours.  This scales linearly with the number of virions in the cell's vicinity.  3) Chosen as a plausible median time (1,000 minutes) between 6 hours and 24 hours.  4) T cells induce apoptosis in nearby virus-secreting epithelial cells at a probabilistic rate such that the expected time to induce apoptosis is 10 minutes.  This rate does not scale with T cell numbers.  5) Calculated for low T cell densities.  6) Chosen to be at the lower end of biologically plausible values because increased T cell counts are shown not to affect the model behavior. }
\label{tab:parameters}
\end{center}
\end{table}
% 7) IgM presence is abstracted by increasing viral decay by a factor of ten. $^\ddagger$Activated T Cell.  $^\ast$Focus of Infection.

>>>>

{\color{dkblue}Identifying the critical components of this complex process is key to understanding viral pathogenicity and designing therapeutic interventions.  However, determining the relative contribution of these myriad experimental components is challenging.  In this study, human epithelial cells were infected with Avian, seasonal, and pandemic influenza \textit{in vitro} and were measured for virus and various cytokine concentrations over the course of 48 hours.  The cytokine profile revealed that the only difference between influenza strains in epithelial cells is in the production of chemokines CXCL10 (IP-10) and CCL5 (RANTES).  Normally chemokines attract T cells to infection sites, and because \textit{in vivo} cell movement in lung tissue cannot be observed directly, we developed a model to examine this effect.}

%We address this challenge with computational modeling, which gives us the advantage of studying relative contributions of different aspects of the host immune response and accounts for strain-specific viral dynamics in the model.

Rather than developing a comprehensive immune system model, we focus on the role of chemokines in directing antigen-specific CD8 T cells from the lymph node to the focus of infection (FOI) in the human lung.  We assume that chemotactic gradients play an important role in guiding T cells to the FOI and that T cells resist extravasation and recirculate in the absence of inflammatory signals.  Chemotactic proteins are known to enhance T cell recruitment in both acute infections and chronic inflammatory diseases \cite{Bromley2008, Medoff2005, Castellino2006 ,Gunn1998, Okada2005}, where recruited T cells mediate either protection or immunopathology.  Infected epithelial cells secrete chemokines \cite{Chan2005}, and contact with CD8 T cells also triggers secretion \cite{Zhao2000}. Activated CD8 T effector cells upregulate chemotactic protein receptors, particularly CXCR3 and CCR5 \cite{Hoji2005, Groom2011a}, responding to their primary ligands IP-10 and RANTES respectively.  Previous work showed that efficient recruitment of T cells to the FOI is crucial for the eventual clearance of the virus \cite{Cerwenka1999, Kim2011}.  We extend this work to examine how the unique properties of the different strains of influenza may affect this process.

Because the recruitment of T cells through chemotaxis is a spatial process, we present a spatial agent-based model (ABM) of the secondary immune response to an influenza infection in the lung.  Previous models have shed light on the details of the primary and secondary immune response, some by applying ordinary differential equation (ODE) models to empirical data \cite{Miao2010, Saenz2010, Handel2008, Lee2009, Murillo2013}, and some by using spatial models to examine the interactions between dendritic cells and T cells in the lymph node \cite{Beltman2007, Zheng2008, Vroomans2012, Textor2014}.  We parameterize our model with \textit{in vitro} generated measurements of viral expansion dynamics as well as published data of host immune responses to flu infection.  The model was used to determine the pathogenic potential of the three individual flu strains: seasonal H1N1 (sH1N1), Avian H5N1 (aH5N1), and pandemic H1N1 (pH1N1).  These three strains were selected for their differences in \textit{in vitro} replication rates \cite{Mitchell2011} as well as differences in \textit{in vivo} severity of human and animal pneumonia and mortality.  The model describes the dynamic, spatially explicit evolution of viral infections in the context of the developing CD8 T cell response.  Sensitivity analysis shows most chosen model parameters have a limited effect on the qualitative results.   Those that have a strong effect are candidates for explaining the dramatic observed differences in pathogenicity among these different flu strains.

@@

{\color{dkblue}The main contributions of this paper are as follows.  First, \textit{in vitro} influenza infection of human epithelial cells stimulated production of chemokines IP-10 and RANTES, suggesting that these two chemokines play a direct role in recruiting T cells to the focus of infection.  Second, the spatial nature of the model allows us to identify physical constraints on T cells' ability to clear the rapidly replicating pH1N1 infection.  Because T cells in the model use the chemokine gradient to locate the FOI, they tend to cluster in areas of high chemokine concentration.  The model suggests that the chemokine response of the host cells may actually hinder T cells' ability to find and clear infected cells in the presence of the rapidly replicating pandemic strain.  Finally, the sensitivity analysis shows the isolated effects of each parameter in the model.  Based on this analysis, each parameter was classified as sensitive or not.  We find that most parameters can be individually varied across a broad range of values without strong effect on predicted outcomes.  Further, the parameter classification is consistent across each broad parameter type. Specifically, chemokine-related parameters are stable within biologically plausible ranges.  In contrast, parameters related to viral kinetics have a strong effect on model results.  We conclude that influenza infection dynamics are dominated by the replication properties of the virus and spatial properties of the infection, confirming that the higher replication rate of 2009 pH1N1 influenza obscured contributions of the chemotactic signals, leading to uncontrolled infection \textit{in silico}.}

@@

The model focuses on activated CD8 T cells migrating through the vascular and lymph networks, including their movement over tissue after extravasation (Fig.~\ref{fig:systemchart}).  {\color{dkblue}While the volume of the lung consists of spherical alveoli, the spread of influenza throughout the lung occurs through the interconnected bronchiole-alveoli complex, which can be viewed as a connected monolayer.  Thus, we represent the lung as a two-dimensional sheet of healthy epithelial cells (S1.3).}   Because we assume activated T cells descend at random into branches of the vascular network, T cells in the model are introduced uniformly at random across the modeled lung surface.  If cytokine signal is detected on the local endothelium the T cell remains in tissue and follows the chemotactic gradient to the FOI.  T cells that do not encounter cytokine recirculate to the lymph node where they reenter the bronchial vascular network.  Once inside the endothelium, T cells follow the chemokine gradient to localized areas of maximum chemokine concentration.  When a T cell contacts an infected epithelial cell it induces apoptosis (S1.4).  Otherwise, infected cells continue to secrete virus for a fixed time and then die.

@@

Our model captures important aspects of T cell response to influenza infection which have not been addressed by earlier T cell models.  The results of our sensitivity analysis are important for future biological modeling and experimental design.  Empirical verification of the model's sensitive parameters (viral response to IgM, infectivity, viral decay rate, viral diffusion rate, and viral production rate) will be valuable to future studies.  Finally, both qualitative and quantitative analysis of our model suggests therapeutic interventions of influenza infection in the lung will likely be most effective by focusing on viral control rather than enhancing T cell responses.

@@

\begin{table}[!ht]
\begin{center}
\begin{tabular}{| l | c | l l l | c |}
  \hline                        
  Referenced Parameters & Units & Min & Value & Max & Source \\
  \hline
  $^{\dagger}$Viral Diffusion in Airway & \textit{$\mu$m$^2$/s} & 3.18e-4 & \textbf{3.18e-2} & 3.18  & \cite{Beauchemin2006} \\
  $^{\dagger}$Viral Decay in Airway & \textit{day$^{-1}$} & 0.01 & \textbf{1} & 100 & \cite{Lee2009} \\
  Chemokine Diffusion Rate & \textit{$\mu$m$^2$/s} & 3.18e-3 & \textbf{.318} & 318 & \cite{Beauchemin2006} \\
  Incubation Time & \textit{hours} & 5 & \textbf{10} & 20 & \cite{Mitchell2011} \\
  Epithelial Cell Radius & \textit{$\mu$m} & | & \textbf{5} & | & \cite{Elbert1999} \\
  T Cell Radius & \textit{$\mu$m} & | & \textbf{5} & | & \cite{abbas2011cellular} \\
  T Cell Production Rate & \textit{cells/h} & 125 & \textbf{1,257} & 3,750 & \cite{Miao2010} \\ 
  T Cell Speed & \textit{$\mu$m/min} & 6e-2 & \textbf{6} & 600 & \cite{Egen2011} \\
  Blood Circulation Time & \textit{seconds} & 1 & \textbf{6} & 3,600 & \cite{Banerjee2010b} \\
  T Cell Sensitivity to Chemokine & \textit{ng/mL} & | & \textbf{100} & | & \cite{Nandagopal2011} \\
  Onset of T Cell Lymph Node Exit & \textit{days }& | & \textbf{5} & | & \cite{Banerjee2011} \\
  $^{\dagger}$IgM Viral Decay Factor & | & 1 & \textbf{10} & 1,000 & \cite{Diamond2003} \\
  IgM Onset & \textit{days} & | & \textbf{4} & | & \cite{Diamond2003} \\
  \hline
  \hline                        
  Estimated Parameters & Units & Min & Value & Max & Footnote \\
  \hline
  Chemokine Decay Rate & \textit{Hz} & 3.8e-6 & \textbf{3.8e-4} & 3.8e-2 & 1\\
  $^{\dagger}$Infectivity & \textit{min/virion} & 12 & \textbf{120} & 1,200   &  2 \\
  Expression Time & \textit{min} & 100 & \textbf{1,000} & 3,000 & \cite{Mitchell2011}$^3$ \\
  T Cell Expected Kill Time & \textit{min} & 0 & \textbf{10} & 100 & 4 \\
  Apoptosis Time & \textit{hours} & 0 & \textbf{1} & 2 & \cite{Ganusov2008}$^5$ \\
  T Cell Age (at FOI) & \textit{min} & 12 & \textbf{120} & 1,200 & 6 \\
  T Cell Age (in Blood) & \textit{days} & 0.04 & \textbf{4} & 400 & 6 \\
  \hline  
\end{tabular}
\caption{Default values used for the model in bold.  Min and max represent the extreme values tested in the sensitivity analysis (Table~\ref{tab:sensitivity} and Fig.~S3-5).  $^{\dagger}$ denotes parameters determined to be sensitive by the sensitivity analysis.  Values were taken from experimental literature if possible and from earlier modeling papers if not.  Parameters not found in literature were estimated as follows.  1) Corresponds to a 30 minute half-life.  2) Epithelial cells are infected at a probabilistic rate such that the expected time for infection in the presence of a single virion is 2 hours.  This scales linearly with the number of virions in the cell's vicinity.  3) Chosen as a plausible median time (1,000 minutes) between 6 hours and 24 hours.  4) T cells induce apoptosis in nearby virus-secreting epithelial cells at a probabilistic rate such that the expected time to induce apoptosis is 10 minutes.  This rate does not scale with T cell numbers.  5) Calculated for low T cell densities.  6) Chosen to be at the lower end of biologically plausible values because increased T cell counts are shown not to affect the model behavior. }
\label{tab:parameters}
\end{center}
\end{table}
% 7) IgM presence is abstracted by increasing viral decay by a factor of ten. $^\ddagger$Activated T Cell.  $^\ast$Focus of Infection.
